Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-04 article.Rmd
text review

Research institutes organised in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this DFG funding program, contributing to the diversity of schemes in Germany. But in response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

@@

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we reviewed the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending to support fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate  Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average Euro foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis for which neither price information nor the journal type could be obtained. In the case of Jisc's 2014 data [@jisc_14], 4.861.772 € being spent on 2,812 publications were not eligible for further analysis because no publication type was given in the data-set.

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data was also added through automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in Crossref.

>>>>

Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this DFG funding program, contributing to the diversity of schemes in Germany. But in response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

@@

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalized publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we reviewed the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending to support fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate  Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average Euro foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis for which either cost information or  journal type was missing. In the case of Jisc's 2014 data [@jisc_14], we could not analyze 4.861.772 € spent on 2,812 publications because no publication type was given in the data-set.

The data collection methods of the Open APC initiative and those of the other initiatives differ in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, in the Wellcome Trust data a publication identifier was added as part of the automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided to only compare German APC spending with that reported by other initiatives. We did not compare the overall size, the distribution over publishers and journal titles, or the indexing coverage in Crossref.
